Rapid technological progress in \\ white light-emitting diodes and its sources in innovation and technology spillovers

### 2023-01-18 2_SSL_EES/article/article_main.tex
updated introduction

\includegraphics{auxiliary/rsc_template_head_foot/dates} & \noindent\normalsize{First commercial white light-emitting diodes (LEDs) were introduced to the market in 1996 . Since then, white LEDs have experienced major improvements in performance and decreases in cost, resulting in rapid market expansion of LED-based solid-state lighting (SSL), one of the key current solutions in global climate change mitigation efforts. Despite the significance of the technology, the extent and sources of these improvements have not been systematically investigated. Understanding what has driven rapid progress in LEDs can provide lessons for accelerating innovation in a broad range of other demand-side clean energy technologies for climate change mitigation. With this aim, we gather systematic evidence on cost and performance improvements in white LEDs from a literature review, interviews with experts in industry and academia, and results from our cost and performance modelling. We find that the overall efficiency of the highest performing warm white LED packages has improved from $\eta_L=5.8\%$ in 2003 to $\eta_L=38.7\%$ in 2020.  During the same period, we estimate that the cost of manufacturing low-power and mid-power LED packages at a U.S. location using state-of-the-art equipment has dropped from $1.1\$$ to $0.05\$$ (in 2020 USD) between 2003 and 2020, a $95.5\%$ decrease. We also show that technology spillovers—i.e., knowledge originating in other technologies—affected all performance dimensions of LEDs. They contributed no less than $8.5\%$ of the total efficiency improvements of devices and were responsible for $\sim 100\%$ of the improvements in consumer experience metrics. The increase in wafer size and manufacturing yield improvements were the primary causes of LED manufacturing cost reductions.}

@@

A rapid reduction of global carbon dioxide emissions is urgently required in order to mitigate the effects of climate change \cite{Forster2019}. According to the United Nations, by the end of 2021, more than 130 countries have set or are considering setting a net zero emissions target by or around 2050 \cite{un2021climate}. The European Union, for example, has set a goal of net zero emissions by 2050—a goal it aims to meet with the help of the European Green Deal, alongside other EU and national policies \cite{eu2020green}.

Achieving these ambitious and critically important targets will require both the deployment of new clean energy technologies, and the acceleration of innovation in existing supply-side \cite{sinn2012green} and demand-side energy technologies \cite{rgeVorsatz2009}. To ensure rapid adoption of these technologies, significant reductions in their costs and improvements in their performance and consumer experience are needed. This requires understanding how these cost reductions and performance improvements can be achieved \cite{Stephan2021}\cite{Ziegler2021}.

There are a range of mechanisms that contribute to improvements in a technology over time, including targeted research and development (R\&D) efforts, economies of scale, learning by doing \cite{Arrow1971}, and economies of scope \cite{johansson2012global}\cite{national2016power}\cite{iea2020perspectives}. The role of these factors at different stages of the innovation life cycle \cite{grubler2012policies}, from research and technology development to demonstration, market formation and diffusion, is an area of active research \cite{Mowery1979} \cite{kavlak2018evaluating} \cite{Ziegler2021}. Innovation is also driven by forces of supply and demand. Various \textit{"technology-push"} drivers reduce the costs of innovating, while \textit{"market-pull"} drivers increase the pay-offs from investing in innovation \cite{anadon2009policy}.

\begin{figure*}[h]
 \centering
 \includegraphics[height=7cm]{2_SSL_EES/article/figures/history_efficacy.pdf}
 \caption{Historical development of the luminous efficacy ($K$) of the most widely-used lighting technologies in human history from 1850 to 2020. Data shown for best performers by year of market introduction, with dashed lines for visual aid. The physical limit for an ideal light source with a colour rendering index of CRI=90, denoted as $K_{max}^{CRI90}$, is shown as a black horizontal line, as per calculations by Murphy et al. \cite{Murphy2012}.  The magnified plot shows the progress in cool white LEDs from 1996 to 2020, with the dashed line indicating a linear rate of efficacy improvement of 10lm/W per year. For comparison, efficacies of best performers in legacy lighting technologies for 2020 are shown as coloured horizontal lines. Note the logarithmic scale of the vertical axis on the main plot and the linear scale on the magnified plot. Abbreviations: HID – High-Intensity Discharge; CFL – Compact Fluorescent Lamp; Hal. – Halogen, Incd. – Incandescent. Source: own synthesis of published data based on a visual approach proposed by Azevedo et al. \cite{azevedo2009transition}. See Supplementary Information, section X, for the full list of data sources. }
 \label{fgr:history_efficecy}
\end{figure*}

The stages and drivers of the innovation life-cycle for a particular technology play out within a broader innovation system \cite{grubler2012policies} \cite{Anadon2016}. Among these drivers, the role of external knowledge transfer and technology spillovers in research and development of energy technologies is an understudied area \cite{Stephan2021}. While the exact definition of spillovers in literature depends on the context and the research question \cite{liu2003does}\cite{Nemet2012}, we follow Stephan et al. \cite{Stephan2021} and consider knowledge to be external – thus a spillover –  if it has been developed for application in other technologies, sectors, or scientific disciplines. There is emerging evidence that understanding spillovers and the knowledge network outside a technology may be an important factor in understanding \cite{Pichler2020} and shaping \cite{Clark2016} \cite{ Stephan2021} \cite{Sun2021} the future evolution of technologies.

Among demand-side technologies, the provision of lighting is a particularly important area for climate change mitigation efforts, as it currently accounts for $15-19\%$ of global electricity consumption \cite{Zissis2016}\cite{doe_electricity}. It is also an area of rapid recent technological change: since the introduction of the first commercial white light-emitting diodes (LEDs) in 1996, lighting technology has experienced dramatic efficiency improvements. As shown in Figure \ref{fgr:history_efficecy}, thanks to the introduction of LED-based solid-state lighting (SSL), the efficacy of lighting sources has increased by three orders of magnitude in just over 20 years, which is significantly faster than the historical progress observed in previous lighting technologies \cite{weinold2021quantifying}. For comparison, the highest performing light-emitting devices today reach efficacies of 220 lm/W \cite{lumistrips2021mid}, while an incandescent light bulb can only reach efficacies of up to 18 lm/W. Moreover, this rapid improvement in efficiency has been accompanied by a similarly impressive decrease in LED manufacturing costs and retail prices. During the same period, LED retail prices have fallen by two orders of magnitude, as shown in Figure \ref{fgr:cost_lamp_small}.

@@

\section*{Acknowledgements}
This research is supported by a grant of the Alfred P. Sloan Foundation titled \href{http://web.archive.org/web/20200623065733/https://sloan.org/grant-detail/8567}{\textit{“What factors drive innovation in energy technologies? The role of technology spillovers and government investment”}}. Michael Weinold further gratefully acknowledges support by the \href{https://www.studienstiftung.ch/}{Swiss Study Foundation}. We would like to thank Venkatesh Narayanamurti, Gabriel Chan, Anna Goldstein, Didier Sornette, and participants of the SPIE West 2021 Conference and \href{https://www.ceenrg.landecon.cam.ac.uk/}{C-EENRG} Seminar Series of the University of Cambridge for many helpful discussions and feedback. We also express our deep gratitude to all interviewees for their willingness to participate in this study and their valuable contributions.

>>>>

\includegraphics{auxiliary/rsc_template_head_foot/dates} & \noindent\normalsize{First commercial white light-emitting diodes (LEDs) were introduced to the market in 1996. Since then, white LEDs have experienced major improvements in performance and decreases in cost, resulting in rapid market expansion of LED-based solid-state lighting (SSL), one of the key current solutions in global climate change mitigation efforts. Despite the significance of the technology, the extent and sources of these improvements have not been systematically investigated. Understanding what has driven rapid progress in LEDs can provide lessons for accelerating innovation in a broad range of other demand-side clean energy technologies for climate change mitigation. With this aim, we gather systematic evidence on cost and performance improvements in white LEDs from a literature review, interviews with experts in industry and academia, and cost and performance modelling. We find that the overall efficiency of the highest performing warm white LED packages has improved from $\eta_L=5.8\%$ in 2003 to $\eta_L=38.7\%$ in 2020. During the same period, we estimate that the cost of manufacturing low-power and mid-power LED packages at a U.S. location using state-of-the-art equipment has dropped from 1.1\$ to 0.05\$ (in 2020 USD) between 2003 and 2020, a 95.5\% decrease. We also show that technology spillovers — i.e., knowledge originating in other technologies—affected all performance dimensions of LEDs, contributing no less than 8.5\% of the total efficiency improvements and $\sim$100\% of the improvements in consumer experience metrics. The increase in wafer size and manufacturing yield improvements were the primary causes of LED manufacturing cost reductions.}

@@

A rapid reduction of global carbon dioxide emissions is urgently required in order to mitigate the effects of climate change \cite{Forster2019}. According to the United Nations, by the end of 2021, more than 130 countries have set or are considering setting a net zero emissions target by or around 2050 \cite{un2021climate}. The European Union, for example, has set a goal of net zero emissions by 2050—a goal it aims to meet with the help of the European Green Deal, alongside other EU and national policies \cite{eu2020green}. The United Kingdom similarly has adopted a similar strategy to achieve net-zero by 2050 \cite{noauthor_ieairena_2023}.

Achieving these ambitious and critically important targets will require both the deployment of new clean energy technologies, and the acceleration of innovation in existing supply-side \cite{sinn2012green} and demand-side energy technologies \cite{rgeVorsatz2009}. To ensure rapid adoption of these technologies, significant reductions in their costs and improvements in their performance and consumer experience are needed. This requires understanding how these cost reductions and performance improvements can be achieved \cite{Stephan2021} \cite{Ziegler2021}.

There are a range of mechanisms that contribute to improvements in a technology over time, including targeted research and development (R\&D) efforts, economies of scale, learning by doing \cite{Arrow1971}, and economies of scope \cite{johansson2012global}\cite{national2016power}\cite{iea2020perspectives}. The role of these factors at different stages of the innovation life cycle \cite{ grubler2012policies}, from research and technology development to demonstration, market formation and diffusion, is an area of active research \cite{Mowery1979} \cite{kavlak2018evaluating} \cite{Ziegler2021}. Innovation is also driven by forces of supply and demand. Various \textit{"technology-push"} drivers reduce the costs of innovating, while \textit{"market-pull"} drivers increase the pay-offs from investing in innovation \cite{anadon2009policy}.

\begin{figure*}[h]
 \centering
 \includegraphics[height=7cm]{2_SSL_EES/article/figures/history_efficacy.pdf}
 \caption{Historical development of the luminous efficacy ($K$) of the most widely-used lighting technologies in human history from 1850 to 2020. Data shown for best performers by year of market introduction, with dashed lines. A multi-order polynomial fit was used to indicate the average rate of change as dashed lines. The physical limit for an ideal light source with a colour rendering index of CRI=90, denoted as $K_max^{CRI90}$, is shown as a black horizontal line, as per calculations by Murphy et al. \cite{Murphy2012}. The magnified plot shows the progress in cool white LEDs from 1996 to 2020, with the dashed line indicating a linear rate of efficacy improvement of 10lm/W per year. For comparison, efficacies of best performers in legacy lighting technologies for 2020 are shown as coloured horizontal lines. Note the logarithmic scale of the vertical axis on the main plot and the linear scale on the magnified plot. Abbreviations: HID – High-Intensity Discharge; CFL – Compact Fluorescent Lamp; Hal. – Halogen, Incd. – Incandescent. Source: own synthesis of published data based on a visual approach proposed by Azevedo et al. \cite{azevedo2009transition}. See Supplementary Information, section , for the full list of data sources. }
 \label{fgr:history_efficacy}
\end{figure*}

The stages and drivers of the innovation life-cycle for a particular technology play out within a broader innovation system \cite{grubler2012policies}\cite{Anadon2016}. Among these drivers, the role of external  knowledge transfer and technology spillovers in research and development of energy technologies is an understudied area  \cite{Stephan2021}. While the exact definition of spillovers in literature depends on the context and the research question \cite{Liu2003}\cite{Nemet2012}, we follow Stephan et al. \cite{Stephan2021} and consider knowledge to be external – thus a spillover –  if it has been developed for application in other technologies, sectors, or scientific disciplines. There is emerging evidence that understanding spillovers and the knowledge network outside a technology may be an important factor in understanding \cite{Pichler2020} and shaping \cite{Clark2016}\cite{Stephan2021}\cite{Sun2021}\cite{kolesnikov_framework_2022} the future evolution of technologies.

Among demand-side technologies, the provision of lighting is a particularly important area for climate change mitigation efforts, as it currently  accounts for 15-19\% of global electricity consumption \cite{Zissis2016}\cite{doe_electricity}. It is also an area of rapid recent technological change: since the introduction of the first commercial white light-emitting diodes (LEDs) in 1996, lighting technology has experienced dramatic efficiency improvements. As shown in Figure \ref{fgr:history_efficacy}, thanks to the introduction of LED-based solid-state lighting (SSL), the efficacy of lighting sources has increased by three orders of magnitude in just over 20 years, which is significantly faster than the historical progress observed in previous lighting technologies \cite{weinold2021quantifying}. For comparison, the highest performing light-emitting devices today reach efficacies of 220 lm/W \cite{lumistrips2021mid}, while an incandescent light bulb can only reach efficacies of up to 18 lm/W. Moreover, this rapid improvement in efficiency has been accompanied by a similarly impressive decrease in LED manufacturing costs and retail prices. During the same period, LED retail prices have fallen by two orders of magnitude, as shown in Figure 2.

@@

\section*{Acknowledgements}
This research is supported by a grant of the Alfred P. Sloan Foundation titled \href{http://web.archive.org/web/20200623065733/https://sloan.org/grant-detail/8567}{\textit{“What factors drive innovation in energy technologies? The role of technology spillovers and government investment”}}. Michael Weinold further gratefully acknowledges support by the \href{https://www.studienstiftung.ch/}{Swiss Study Foundation}. We would like to thank Venkatesh Narayanamurti, Gabriel Chan, Anna Goldstein, Didier Sornette, and participants of the SPIE West 2021 Conference and \href{https://www.ceenrg.landecon.cam.ac.uk/}{C-EENRG} Seminar Series of the University of Cambridge for many helpful discussions and feedback. We also express our deep gratitude to all interviewees for their willingness to participate in this study and their valuable contributions. Support by German consumer organization \textit{Stiftung Warentest} in providing access to their archive is gratefully acknowledged.
